---
title: "Data Science Walkthrough"
author: "Brendan Cron, Everett Brown, Samantha Hilbert"
date: "5/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Welcome to the world of data science! This webpage will give you a walkthrough of some of the core concepts. This project is written in the language of R but should be explained in enough detail so someone that has never seen R can easily follow. 
The data we will be monitoring is coronavirus and how it has affected different aspects of our lives. We will look at everthing from dates that states start quarentine to how many people google searched how to make sourdough bread. Hope you enjoy!

## Data Parsing

The first part of data science is collecting data. This can be done manually with surveys but it is much more efficient to look for an existing dataset on the internet. There are many ways that we can get the data, 2 of those ways that we will look at is using a comma seperated value (CSV) file or through webscraping.
Lets first look at how to get data from a CSv. For this example we grabbed a dataset from the European Centre for Disease Prevention and Control. This csv is in this working directory so we can directly import it to our project

```{r covidCSV}
covidData <- read.csv(file = 'covid.csv')
covidData
```